### 2018-08-03 conclusion.tex
Final improvements

The pipelines presented in chapter \ref{chap:pipeline} offer an output-rich and easy to deploy cost-free pieces of software. All of them have pros and cons, and their combined usage was found to provide the best results, underlining the importance of the development of programs implementing open data formats exchangeable across tools. More functionalities can be added, including \textit{de novo} and \ac{PTM} search, fractioned samples and missing data management. BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model}, can be passed any peptide level dataset and provide estimates of the uncertainty beyond frequentist \ac{log2FC} estimates. It can be customized to expand its currently implemented model. The lack of success in the attempt to model the peptide bias using the aminoacidic sequence reveals the complexity of its origin. Recently developed tools  could be applied in future work to extract a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach. The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software, and it is encouraged to continue on that line.

>>>>

The pipelines presented in chapter \ref{chap:pipeline} offer an output-rich and easy to deploy cost-free pieces of software. All of them have pros and cons, and their combined usage was found to provide the best results, underlining the importance of the development of programs implementing open data formats exchangeable across tools. More functionalities can be added, including \textit{de novo} and \ac{PTM} search, fractioned samples and missing data management. BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model}, can be passed any peptide level dataset and provide estimates of the uncertainty beyond frequentist \ac{log2FC} estimates. It can be customized to expand its currently implemented model. The lack of success in the attempt to model the peptide bias using the aminoacidic sequence reveals the complexity of its origin. Recently developed tools  could be applied in future work to extract a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach. The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software. In summary, this work grants exciting opportunities for developing workflows in the field.

### 2018-08-03 introduction.tex
Backup and spell check
Final improvements

Novozymes A/S \ac{NZ} is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes. The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact \cite{madigan2000brock}. Notorious examples of such applications include waste-water treatment, household care or fermentation. For example, a product called Fermax was recently released to prevent foam development during the sugarcane ethanol production (see figure \ref{fig:ethanol}).

@@

The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}. Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans \cite{Nelson2008}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
  \caption[The 20 aminoacids]{Molecular structure of the 20 aminoacids, organized by chemical properties into five groups. Taken from \footnote{\href{https://cnx.org/contents/vxf032Bc@4/Proteins}{https://cnx.org/contents/vxf032Bc@4/Proteins}}.}
  \label{fig:aminoacids}
\end{figure}

>>>>

Novozymes A/S (\ac{NZ}) is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes. The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact \cite{madigan2000brock}. Notorious examples of such applications include waste-water treatment, household care or fermentation. For example, a product called Fermax was recently released to prevent foam development during the sugarcane ethanol production (see figure \ref{fig:ethanol}).

@@

The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}. Enzymes consist of an extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans \cite{Nelson2008}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
  \caption[The 20 aminoacids]{Molecular structure of the 20 aminoacids, organized by chemical properties into five groups. Taken from \href{https://cnx.org/contents/vxf032Bc@4/Proteins}{https://cnx.org/contents/vxf032Bc@4/Proteins}.}
  \label{fig:aminoacids}
\end{figure}

### 2018-08-03 modelling.tex
Final improvements

The current label-free quantification methods, presented in chapter \ref{chap:pipeline}, rely on frequentist statistics, which return point estimates of model parameters such as the abundance ratio (fold change) across conditions. However, a Bayesian  approach to this problem, that we know of, is lacking in the literature. As a response to this shortcoming, a statistical model, named BayesQuant, was written in the probabilistic programming framework PyMC3 and tested on the proteome benchmark dataset. The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.

@@

The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}. This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a \textit{p-value}, which measures the probability of the null hypothesis. The null hypothesis usually states that the true value of the parameter is 0. Thus, \textit{p-values} do not say anything about the probability of hypotheses outside $H_0$ and $H_1$.

@@

It can be applied to the quantification problem, where $\theta$ turns into the \ac{log2FC} parameter we try to estimate. All is needed is the computation of the posterior probability distribution. Unfortunately, this is tractable analytically for simple models only. However, Markov Chain-Monte Carlo (\ac{MCMC}) and Variational Inference (\ac{VI}) methods can be used to approximate this posterior. Both of which are now possible to run thanks to the advent of modern computing.

@@

The study of the \ac{log2FC} posterior probability distribution (see figure \ref{fig:posteriors}) shows the most likely parameter values and empowers Bayesian Null Hypothesis Testing. The 95\% High Probability Density Interval (\ac{HPDI}), which is the narrowest interval containing 95\% of the total probability, indicates the most likely true values of the parameter. Presence or absence of overlap between the 95\% \ac{HPDI} and the Region Of Practical Equivalence (\ac{ROPE}) provides a formal way of discarding a point value of the parameter. The \ac{ROPE} is a small interval considered to be essentially the same as the null value \cite{Kruschke}. In this and all future applications in the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.

@@

In order to further validate BayesQuant\textquotesingle s performance, it was tested on groups of 5 proteins, one group from each proteome. Each pair of groups was made by proteins with a different number of observed peptides (2, 3, 4, 6, 7, 10) (see figure \ref{fig:dumbbell}). The more observed peptides, the more accurate the estimation will be as more data will be available. Since the \ac{HPDI} reflects the uncertainty, a decrease on its width was expected to be observed (see figure \ref{fig:dumbbell}).

>>>>

The current label-free quantification methods, presented in chapter \ref{chap:pipeline}, rely on frequentist statistics, which return point estimates of model parameters such as the abundance ratio (fold change) across conditions. However, a Bayesian approach to this problem, that we know of, is lacking in the literature. As a response to this shortcoming, the thesis focused on writing a statistical model, named BayesQuant, in the probabilistic programming framework PyMC3 and testing it on the proteome benchmark dataset. The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.

@@

The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}. This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a \textit{p-value}, which measures the probability of the null hypothesis. The null hypothesis usually states that the true value of the parameter is 0. Thus, \textit{p-values} do not say anything about the probability of hypotheses outside $H_0$ and $H_1$. Is there a framework that reports results for a particular hypothesis, based just on the seen data and our prior knowledge?

@@

It can be applied to the quantification problem, where $\theta$ turns into the \ac{log2FC} parameter we try to estimate. All is needed is the computation of the posterior probability distribution. Even though, this is tractable analytically for simple models only, fortunately Markov Chain-Monte Carlo (\ac{MCMC}) and Variational Inference (\ac{VI}) methods can be used to approximate this posterior when the analytical derivation is not possible. Both \ac{MCMC} and \ac{VI} are now possible to run thanks to the advent of modern computing.

@@

The study of the \ac{log2FC} posterior probability distribution (see figure \ref{fig:posteriors}) shows the most likely parameter values and empowers Bayesian Null Hypothesis Testing. The 95\% High Probability Density Interval (\ac{HPDI}), which is the narrowest interval containing 95\% of the total probability, indicates the most likely true values of the parameter. Presence or absence of overlap between the 95\% \ac{HPDI} and the Region Of Practical Equivalence (\ac{ROPE}) provides a formal way of discarding a point value of the parameter. The \ac{ROPE} is a small interval considered to be essentially the same as the null value \cite{Kruschke}. In this and all future applications of the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.

@@

Could these results be extended to other proteins in the dataset and validate the tool? Six groups of 5 proteins, with a different number of observed peptides (2, 3, 4, 6, 7, 10) were organized in each proteome and employed to test BayesQuant (see figure \ref{fig:dumbbell}). The more observed peptides, the more accurate the estimation is expected to be as more data for the same protein becomes available. Since the \ac{HPDI} reflects the uncertainty, a decrease on its width is thus expected to be observed (see figure \ref{fig:dumbbell}).

### 2018-08-03 msreview.tex
Final improvements

The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide like as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}. It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}) \cite{Barsnes2008}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{mass_spectrometer}
\caption[Mass spectrometer diagram]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
\label{fig:mass_spectrometer}
\end{figure}

\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
\stepcounter{footnote}

@@

The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}). These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.

>>>>

The mass spectrometer is the ensemble of pieces of equipment analyzing peptides like those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}. It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}) \cite{Barsnes2008}.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{mass_spectrometer}
\caption[Mass spectrometer diagram]{Schematic view of a mass spectrometer. Taken from \href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}.}
\label{fig:mass_spectrometer}
\end{figure}
%
%\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
%\stepcounter{footnote}

@@

The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly Time Of Flight (\ac{TOF}), Ion Trap (\ac{IT}) and Quadrupole (\ac{Q}). These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.

### 2018-08-03 pipeline.tex
Final improvements

Relative quantification was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob). Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function. In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.

@@

The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) across conditions H and L. It is important to take into account as many of the effects influencing protein quantification as possible. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

>>>>

Relative quantification i.e the computation of $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob). Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function. In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.

@@

Could the apex MS1 intensities extracted by moFF be used as proxy for peptide abundance and perform relative quantification? It is important to take into account as many of the effects influencing the recorded data as possible. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

### 2018-08-03 thp1.tex
Final improvements

Data on 6444 peptides, as collected in the peptide summary file from moFF, was passed to MSqRob, resulting in the quantification of 1433 proteins. A histogram of the \ac{log2FC} reveals that in most cases, not enough evidence was collected to declare the \ac{log2FC} different from 0 (see figure \ref{fig:compomics_rob} A). Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which a non-null estimate was provided were more abundant in the +LPS condition. A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria ($q-value < 0.05$), displayed an absolute value of the \ac{log2FC} less than 1. However, all were considered differentially abundant. The 32 protein groups composed by a single protein (single protein \ac{DAP}s) are shown in table \ref{tab:thp1_rob_results}.

@@

A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background. The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.

@@

These results can be better visualized by making use of a treemap (see figure \ref{fig:treemap}), which makes the interpretation of table \ref{tab:gsea} more straightforward.

@@

The UniprotKB to Gene name mapping of the set of 49 \ac{DAP}s (including multi-protein groups) returned by MSqRob produced a list of 51 entities. The list was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were over-represented compared to the human background. 50 entities were mapped to the database. In agreement with the \ac{GSEA} results, a handful of pathways linked to the immune response were found to be significantly enriched (see table \ref{tab:pathway_enrichment}) (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration. Moreover, the finding of several RNA related pathways exposed the presence of potential cell reprogramming, essential in any change in protein profile.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{string_normal_image}
\caption[Protein interaction network]{\textbf{STRING interaction network}. Every node represents a protein in the list, and an edge between any pair represents evidence for some kind of association or interaction between them. Evidence is either experimentally determined or predicted using diverse methods. Red colored nodes are included in the "immune system process" set.}
\label{fig:STRING}
\end{figure}

@@

One of the most frequent issues found when analyzing proteomics datasets is the handling of missing values \cite{Lazar2016}. The presence of missing data in many of the rows in the data passed to BayesQuant led to the dropping of 4803 out of 6443 peptides, producing a peptide dataset of only 1641 data points that accounted for 350 protein groups. Thanks to the missing data handling implemented in MSqRob, this tool quantified 1433 proteins, enough to continue the quantitative analysis (see table \ref{tab:balance}). However, the fact that the number of observed groups was 2787 implies that the management of missing values could potentially be improved in both tools, and particularly in BayesQuant.

@@

Missing data should be imputed at the peptide level. The implementation of a module performing this task prior to quantification would  improve the accuracy and throughput of the quantification process. Several approaches have been proposed, some within a probabilistic framework \cite{Lazar2016}. For example, in cases of clear \ac{MNAR}, imputation could be based on an inferred global \ac{LOD}, so that missing values are replaced with it. Work on this line could potentially boost the performance of BayesQuant.

>>>>

Data on 6444 peptides, as collected in the peptide summary file from moFF, was passed to MSqRob, resulting in the quantification of 1433 proteins. A histogram of the \ac{log2FC} reveals that in most cases, not enough evidence was collected to declare the \ac{log2FC} different from 0 (see figure \ref{fig:compomics_rob} A). Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which a non-null estimate was provided were more abundant in the +LPS condition. A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria (\textit{q-value} $< 0.05$), displayed an absolute value of the \ac{log2FC} less than 1. However, all were considered differentially abundant. The 32 protein groups composed by a single protein (single protein \ac{DAP}s) are shown in table \ref{tab:thp1_rob_results}.

@@

A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) with the gProfileR \cite{gprofiler} illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background. The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.

@@

A Revigo \cite{Supek2011} treemap (see figure \ref{fig:treemap}) makes the interpretation of table \ref{tab:gsea} more straightforward.

@@

The UniprotKB to Gene name mapping of the set of 49 \ac{DAP}s (including multi-protein groups) returned by MSqRob produced a list of 51 entities. The list was supplied to the ComPath Pathway Enrichment tool \cite{Domingo2018} to highlight which cellular pathways were over-represented compared to the human background. 50 entities were mapped to the database. In agreement with the \ac{GSEA} results, a handful of pathways linked to the immune response were found to be significantly enriched (see table \ref{tab:pathway_enrichment}) (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration. Moreover, the finding of several RNA related pathways exposed the presence of potential cell reprogramming, essential in any change in protein profile.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{string_normal_image_confidence}
\caption[Protein interaction network]{\textbf{STRING interaction network}. Every node represents a protein in the list, and an edge between any pair represents evidence for some kind of association or interaction between them. Evidence is either experimentally determined or predicted using diverse methods. The thickness of the edges represents the level of confidence in that interaction. Red colored nodes are included in the "immune system process" set.}
\label{fig:STRING}
\end{figure}

@@

One of the most frequent issues found when analyzing proteomics datasets is the handling of missing values \cite{Lazar2016}. The presence of missing datapoints led BayesQuant to drop 4803 out of 6443 peptides, producing a peptide dataset of only 1641 rows that accounted for 350 protein groups. Thanks to the missing data handling implemented in MSqRob, the latter quantified 1433 proteins, enough to continue the quantitative analysis (see table \ref{tab:balance}). However, the fact that the number of observed groups in the original dataset was 2787 implies that the management of missing values could potentially be improved in both tools, and particularly in BayesQuant.

@@

Missing data should be imputed at the peptide level. Several approaches have been proposed, some within a probabilistic framework \cite{Lazar2016}. For example, in cases of clear \ac{MNAR}, imputation could be based on an inferred global \ac{LOD}, so that missing values are replaced with it. Work on this line could potentially boost the performance of BayesQuant.
